---
title: "chapter6.Rmd"
author: "Laura_Hakkinen"
date: "2023-12-11"
output: html_document
---

# Chapter 6: Analysis of longitudinal data

## PART I

### 6.1 load data and packages

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggplot2)
library(lme4)

setwd("\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\KURSSIT\\Open_Data_Science\\IODS-project\\data")

RATSL <- read.csv2("RATSL.csv")
BPRSL <- read.csv2("BPRSL.csv")

str(RATSL)
str(BPRSL)
```
```{r}

## Convert the categorical variables of both data sets to factors

# Factor Group & ID
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# Factor variables ID and Group
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)

str(RATSL)
str(BPRSL)
```




### 6.2 view of the RATS data
```{r}
names(RATSL)

str(RATSL)

summary(RATSL)


```


### 6.3 Graphical displays of longitudinal data:


```{r}

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = Group, fill=ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

Weight grows in time

Group 1 is quite separate from group 2. and 3.

Lower/higher weight at the beginning indicates lower/higher weight at the end


### 6.4 Standardise

```{r}

# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight)) / sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised Weight
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = Group, fill=ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  scale_y_continuous(name = "standardized Weight")

```


### 6.5 Summary graphs


```{r}
# how many per observations per ID
table(RATSL$ID)

# Number of IDs (per group):
n <- 11

# Summary data with mean and standard error of Weight by Group and Time 
RATSL_S <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL_S)

# Plot the mean profiles
ggplot(RATSL_S, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")


```
Very small standard error (se) in group 1, small se in group 3, and large se in group 2

Mean values show less change in group 1 over time compared to group 2 and 3


### 6.6 Find outliers


```{r}
# time points
unique(RATSL$Time)

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 1)
RATSL10S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL10S)

# Draw a boxplot of the mean versus Group
ggplot(RATSL10S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), time 8-64")

# Create a new data by filtering outliers and adjust the ggplot code to draw the plot again with the new data
RATSL10S1 <- filter(RATSL10S, mean>250)
RATSL10S1 <- filter(RATSL10S1, mean<550)
RATSL10S1 <- filter(RATSL10S1, mean!=495.2)

# Draw a boxplot of the mean versus Group
ggplot(RATSL10S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), time 8-64")


```

After removing outliers the among group variation is very small

### 6.7 T test and Anova

We cannot do T test with 3 groups! So let's just do linear model and anova. But note that this isn't a fibal analysis as it doesn't take i nto account the longitudial form of the data

```{r}

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ Time + Group, data = RATSL_S)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

Both time and group are significant



# PART II: Linear Mixed Effects Models


### 6.8 Plot first


```{r}
dim(BPRSL)

# Plot the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = subject, linetype = treatment))  +
  facet_grid(. ~ treatment, labeller = label_both) +
  geom_line(aes(linetype = treatment)) + scale_x_continuous(name = "week") + scale_y_continuous(name = "bprs") + theme(legend.position = "top")

# group = treatment



```


### 6.9 Linear model

Although, doesn't really apply to longitudial data set!

```{r}
# create a regression model RATS_reg
BPRS_reg <- BPRSL %>% lm(bprs ~ week + treatment, data = .)

# print out a summary of the model
summary(BPRS_reg)
```

Week (and intercept) is significant, but the treatment isn't


###6.10 The Random Intercept Model


```{r}

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```
Quite large estimated variance of random effects (subject) indicating variation in the intercepts

But let's compare to random intercept and random slope model to see if we need randon slope as well


### 6.11 Random Intercept and Random Slope Model


```{r}

# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

AIC is smaller with random slope and p value also significant indicating we should include the random slope in the model


### 6.12 Random Intercept and Random Slope Model with interaction

Finally, we can fit a random intercept and slope model that allows for a treatment × week interaction.


```{r}

# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

```


The model with interaction has a little bit smaller AIC, butu since the difference is small (~1) and the p value is not significant lets use the more parsimonous model **without** interaction


#### Draw the plot of BPRSL with the observed bprs values

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week") +
  scale_y_continuous(name = "Observed bprs") +
  theme(legend.position = "top")
```


#### Plot fitted values

Note! Based on the choces model without interaction!

```{r}
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>% mutate(fitted = Fitted)

ggplot(BPRSL, aes(x = week, y = fitted, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week") +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")


```


Fitted values show that the treatment effect is very small! Could be argued that the difference between the treatments is meaningless, and the bprs score is mostly affected by the random effect of the individual (subject) and time (week)





